#Profile
###Title: "single R markdown"
###Author: "remydozie"
###Date: "23 February 2016"
###Output: html_document

#Introduction
This markdown document shows the steps as well as R Codes and Outputs for data analysis on personal activity monitoring device.This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.
## Data Variables
###steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
###date: The date on which the measurement was taken in YYYY-MM-DD format
###interval: Identifier for the 5-minute interval in which measurement was taken
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.
# Analysis
##Loading and preprocessing the data
##load required libraries
```{r,echo = TRUE}
library(dplyr)
library(lattice)
```
## Reading Data
```{r,echo = TRUE}
#reading the activity data and pre-processing. Make sure data is downloaded and in the right directory
setwd("C:/Users/nnadozie/Documents/GitHub/RepData_PeerAssessment1")
activity <- read.csv('activity.csv', header = TRUE, sep = ",")
activity$date <- as.Date(activity$date)
```
##Question:What is mean total number of steps taken per day?
```{r, echo = TRUE}
###For this part of the assignment, you can ignore the missing values in the dataset.
###Calculate the total number of steps taken per day
activity2 <- group_by(activity,date)
activity2 <- na.omit(activity2)
dailysteps <- summarise(activity2,total_steps = sum(steps))
head(dailysteps)
```
```{r, echo = TRUE}
###Make a histogram of the total number of steps taken each day
#png("plot1.png",width = 480, height = 480)
hist(dailysteps$total_steps,main = "Total steps taken per day",
     xlab = "Total number of steps",col = "blue",
     breaks= length(dailysteps$total_steps)-1,plot = TRUE)
```

```{r, echo = TRUE}
dev.off()
```

```{r, echo = TRUE}
###Calculate and report the mean and median of the total number of steps taken per day
MeanStepsPerDay <- mean(dailysteps$total_steps)
MedianStepsPerDay <-median(dailysteps$total_steps)
```

```{r, echo = TRUE}
MeanStepsPerDay
```

```{r, echo = TRUE}
MedianStepsPerDay
```
## Question: What is the average daily activity pattern?
###1.Make a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

###2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo = TRUE}
activity3 <- group_by(activity, interval)
activity3 <- na.omit(activity3)
avgsteps_interval <- summarise(activity3, avg_steps = mean(steps)) 
```

```{r, echo = TRUE}
#png("plot2.png",width = 480, height = 480)
plot(avgsteps_interval,type = 'l', col = 'brown',lwd=2,
     main = "Average number of steps taken across all days in 5min intervals",
     xlab = "5min intervals",ylab = "average number of steps taken")
```

```{r, echo = TRUE}
dev.off()
```
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo = TRUE}
maxsteps <- filter(avgsteps_interval, avg_steps == max(avg_steps))
cat(sprintf("At %sth 5min interval, the maximum number of %f steps are found on average on all days in the data set. \n", maxsteps$interval, maxsteps$avg_steps))
```
## Managing Missing Values
###Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r, echo = TRUE}
missing_steps <- sum(is.na(activity$steps))
missing_date <- sum(is.na(activity$date))
missing_interval <- sum(is.na(activity$interval))

cat(sprintf("Total number of missing values in steps field =  %s\n", missing_steps))
```


```{r, echo = TRUE}
cat(sprintf("Total number of missing values in date field =  %s\n", missing_date))
```


```{r, echo = TRUE}
cat(sprintf("Total number of missing values in interval field =  %s\n",missing_interval))
```
```{r, echo = TRUE}
#create new dataset with removed NA
activity4<- activity
avgsteps <- summarise(activity3, mean_steps = mean(steps), median_steps = median(steps))
activity4$steps[is.na(activity4$steps)] <- tapply(activity4$steps, activity4$interval, mean, na.rm = TRUE)
```

```{r, echo = TRUE}
#check for any missing values 
sum(is.na(activity4))
```

```{r, echo = TRUE}
activity5 <- group_by(activity4, date)
dailysteps_afms <- summarise(activity5,total_steps = sum(steps))

#png("plot3.png",width = 480, height = 480)
hist(dailysteps_afms$total_steps,breaks=length(dailysteps_afms$total_steps)-1, 
     col = 'red', main = "Histogram of total number of steps taken each day",
     xlab = "Total number of Steps",plot = T)
```
```{r, echo = TRUE}
dev.off()
```

```{r, echo = TRUE}
new_mean <-mean(dailysteps_afms$total_steps)
new_median <-median(dailysteps_afms$total_steps)

cat(sprintf("The new mean is=  %s\n", new_mean))
```

```{r, echo = TRUE}
cat(sprintf("The new median is =  %s\n", new_median))
```
##Question:Are there differences in activity patterns between weekdays and weekends?
For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.
```{r, echo = TRUE}
###Create a new factor variable in the dataset with two levels - “weekday” and “weekend” indicating whether a given date is a weekday or weekend day
###Make a panel plot containing a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
activity6<- activity5
activity6$wdays <- weekdays(as.Date(activity6$date))
activity6 <-mutate(activity6, wk.factor = factor((wdays == "Sunday" | wdays == "Saturday"),levels = c(FALSE,TRUE),
                                     labels = c("weekdays","weekends")))
MeanStep_byIntervalByWkday <- aggregate(steps ~ interval + wk.factor, data=activity6, FUN="mean")
```

```{r, echo = TRUE}
#png("plot4.png",width = 480, height = 480)
xyplot(steps ~ interval | wk.factor, data=MeanStep_byIntervalByWkday, type="l", grid=T, layout=c(1,2),
       main="Plot for comparision of number of steps on weekdays and weekends",
       xlab = "5min. intervals", ylab = "Number of steps",lwd = 2,col = "blue")
```